Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-28 soc_squeezing/soc_squeezing_notes.tex
more parameter tweaking for the squeezing protocols, but this time using the depth exclusion plots

\begin{table}[h]
  \centering
  \caption{Example parameters for spin squeezing via one-axis
    twisting.}
  \label{tab:parameters}
  \begin{tabular}{|l|c|l|}
    \hline
    Parameter & Symbol & Value \\ \hline\hline
    Primary lattice depth & $V_0$ & 5 $E_R$ \\
    Transverse lattice depths & $V_T$ & 50 $E_R$ \\
    SOC strength & $\phi$ & $\pi/50$ \\
    Lattice sites & $L$ & 100 \\
    Atom number & $N$ & 100 \\ \hline\hline
    Tunneling rate & $J_0$ & $\approx230\times2\pi$ Hz \\
    On-site interaction strength & $U$ & $\approx1.3\times2\pi$ kHz \\
    Twisting strength & $\chi$ & $\approx3.2\times2\pi$ mHz \\
    Optimal OAT squeezing time & $t_{\t{opt}}^{\t{OAT}}$
    & $\approx2.5$ seconds \\ \hline
  \end{tabular}
\end{table}

@@

We can go beyond the $\sim1/N^{2/3}$ one-axis twisting noise floor and
achieve an improved $\sim1/N$ noise floor via so-called two-axis
twisting, which can be achieved with external driving protocols.  By
turning on a clock laser resonant on the electronic transition of the
atoms, we can realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = -\Omega \sum_n s_n^x + H_{\t{OAT}}^z
  = -\Omega S_x + \chi S_z^2,
\end{align}
where the drive strength $\Omega$ may generally depend on time, but
must always be much smaller in magnitude than the collective spin gap
$\eta U$ in order preserve the validity of the effective one-axis
twisting Hamiltonian (see Appendix \ref{sec:squeezing_derivation}).
Following the prescription in ref.~\cite{huang2015twoaxis} we now
modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$ with
$\omega\gg N\chi$, and move into the rotating frame of
$-\Omega\p{t}S_x$.  After a single secular approximation which relies
only on $\omega\gg N\chi$, this procedure results in the effective
Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift.  Given
the set of parameters in Table \ref{tab:parameters}, the two-axis
twisting Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive
frequency $\omega=\abs{\eta U\chi}^{1/2}\approx20\times2\pi$ Hz and
modulation index $\beta\approx0.906$, and has an optimal squeezing
time of $t_{\t{opt}}^{\t{TAT}}\approx3.2$ seconds.  The figure of
merit for squeezing is the normalized minimal variance of spin in the
plane orthogonal to the mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.  A comparison of this squeezing parameter
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison_TAT}.

@@

It may be possible to further enhance the twisting strength $\chi$ by
placing the atoms in an optical cavity, as in
ref.~\cite{hu2017vacuum}.  Both squeezing times and twisting strengths
might also be further enhanced by using a 2-D lattice, although more
benchmarking is necessary to verify the validity of the collective
spin model in 2-D.  As shown in Figure
\ref{fig:squeezing_comparison_2D},

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_2D.pdf}
  \caption{Squeezing via one- and two-axis twisting in a 2-D lattice
    with $100\times100$ sites at unit filling, with the remaining
    lattice/SOC parameters given in Table \ref{tab:parameters}.}
  \label{fig:squeezing_comparison_2D}
\end{figure}

>>>>

\begin{table}[h]
  \centering
  \caption{Example parameters for spin squeezing via one-axis
    twisting.}
  \label{tab:parameters}
  \begin{tabular}{|l|c|l|}
    \hline
    Parameter & Symbol & Value \\ \hline\hline
    Primary lattice depth & $V_0$ & 5 $E_R$ \\
    Transverse lattice depths & $V_T$ & 60 $E_R$ \\
    SOC strength & $\phi$ & $\pi/25$ \\
    Lattice sites & $L$ & 100 \\
    Atom number & $N$ & 100 \\ \hline\hline
    Tunneling rate & $J_0$ & $\approx230\times2\pi$ Hz \\
    On-site interaction strength & $U$ & $\approx1.5\times2\pi$ kHz \\
    Twisting strength & $\chi$ & $\approx11\times2\pi$ mHz \\
    Optimal OAT squeezing time & $t_{\t{opt}}^{\t{OAT}}$
    & $\approx0.70$ seconds \\ \hline
  \end{tabular}
\end{table}

@@

We can go beyond the $\sim1/N^{2/3}$ one-axis twisting noise floor and
achieve an improved $\sim1/N$ noise floor via so-called two-axis
twisting, which can be achieved with external driving protocols.  By
turning on a clock laser resonant on the electronic transition of the
atoms, we can realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = -\Omega \sum_n s_n^x + H_{\t{OAT}}^z
  = -\Omega S_x + \chi S_z^2,
\end{align}
where the drive strength $\Omega$ may generally depend on time, but
must always be much smaller in magnitude than the collective spin gap
$\eta U$ in order preserve the validity of the effective one-axis
twisting Hamiltonian (see Appendix \ref{sec:squeezing_derivation}).
Following the prescription in ref.~\cite{huang2015twoaxis} we now
modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$ with
$\omega\gg N\chi$, and move into the rotating frame of
$-\Omega\p{t}S_x$.  After a single secular approximation which relies
only on $\omega\gg N\chi$, this procedure results in the effective
Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift.  Given
the set of parameters in Table \ref{tab:parameters}, the two-axis
twisting Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with a drive
frequency $\omega=\abs{\eta U\chi}^{1/2}\approx41\times2\pi$ Hz and
modulation index $\beta\approx0.906$, and has an optimal squeezing
time of $t_{\t{opt}}^{\t{TAT}}\approx0.89$ seconds.  The figure of
merit for squeezing is the normalized minimal variance of spin in the
plane orthogonal to the mean spin vector $\bk{\v S}$:
\begin{align}
  \xi^2 \equiv \f{N}{\abs{\bk{\v S}}^2}
  \min_{\hat{\v n}\perp\bk{\v S}} \bk{\p{\v S\c\hat{\v n}}^2}
\end{align}
for $\abs{\hat{\v n}}=1$.  A comparison of this squeezing parameter
through $H_{\t{OAT}}^z$ and $H_{\t{TAT}}^{z,x}$ for the initial state
$\ket{Y}\equiv\bigotimes_n\p{\ket{\g}_n+i\ket{\e}_n}/\sqrt2$ via the
parameters in Table \ref{tab:parameters} is provided in Figure
\ref{fig:squeezing_comparison_TAT}.

@@

It may be possible to further enhance the twisting strength $\chi$ by
placing the atoms in an optical cavity, as in
ref.~\cite{hu2017vacuum}.  Both squeezing times and twisting strengths
might also be further enhanced by using a 2-D lattice, although more
benchmarking is necessary to verify the validity of the collective
spin model in 2-D.  Figure \ref{fig:squeezing_comparison_2D} shows a
comparison of squeezing strengths via one- and two-axis squeezing in a
2-D lattice with $100\times100$ sites.

\begin{figure}
  \centering
  \includegraphics{squeezing_comparison_2D.pdf}
  \caption{Squeezing via one- and two-axis twisting in a $V_0=7E_R$
    depth 2-D lattice with $100\times100$ sites at unit filling
    ($N=10^4$), transverse confinement depth $V_T=80E_R$, and
    spin-orbit coupling strength $\phi=\pi/25$.}
  \label{fig:squeezing_comparison_2D}
\end{figure}
